---
title: "Simple test model with and without look-up tables in Stan"
format: html
editor: visual
---

We consider a test problem involving a fixed non-linear transformation $V_x$, $$y_i = V_x(a+bx_i)+\sigma \xi_i,\\
\xi_i \sim N(0,1)\; iid,\\
V_x(x)=0.2x+x^5,$$ where $V_x$ is treated as a known empirical function that will be implemented using a look-up table. We start by generating some tabular data.

```{r}
require(tidyverse)
Vfun <- \(x)(0.2*x+x^5)
Vx_data <- tibble( x=seq(-2,2,0.1) ) |> mutate(Vx=Vfun(x))

```

Next, we choose values for the fit parameters $a,b,\sigma$ and generate some synthetic data.

```{r}
a <- 0.2
b <- 1.5
sig <- 0.25

set.seed(123) # reproducible random numbers
xi <- runif(30)
yi <- Vfun(a+b*xi)+sig*rnorm(length(xi))
train_data <- tibble(  x=xi, yi=yi)


tibble( type="truth", x=seq(0,1,0.01)) |> 
  mutate( Vx=Vfun(a+b*x) ) |> 
  ggplot(aes(x=x, y=Vx, color=type))+
  geom_line()+
  geom_point(data = train_data |> rename(Vx=yi) |> mutate(type="data"))+
  labs( y="y", color="")
```

Next, we fit the model with cmdstan using

1.  the true functional form for $V_x(x)$ in the Stan model file, and
2.  a spline lookup table built on tabulated data, and
3.  a pchip look-up table built on tabulated data.

We use somewhat long sampling partly to test reproducibility and run times..

```{r}
library(cmdstanr)
require(posterior)

stan_data1 <- list(
  N = nrow(train_data),
  x = train_data$x,
  y = train_data$yi
)

stan_data2 <- list(
  N = nrow(train_data),
  x = train_data$x,
  y = train_data$yi,
  K = nrow(Vx_data),
  xk = Vx_data$x,
  Vk = Vx_data$Vx
)

mod1 <- cmdstan_model("Vx_true_function.stan")
fit1 <- mod1$sample(
  data = stan_data1,
  chains = 4,
  parallel_chains = 4,
  iter_warmup = 5000,
  iter_sampling = 20000,
  thin=20,
  seed = 123
)

mod2 <- cmdstan_model("Vx_spline_lookup.stan")
fit2 <- mod2$sample(
  data = stan_data2,
  chains = 4,
  parallel_chains = 4,
  iter_warmup = 5000,
  iter_sampling = 20000,
  thin=20,
  seed = 123
)

mod3 <- cmdstan_model("Vx_pchip_lookup.stan")
fit3 <- mod3$sample(
  data = stan_data2,
  chains = 4,
  parallel_chains = 4,
  iter_warmup = 5000,
  iter_sampling = 20000,
  thin=20,
  seed = 123
)

```

The results are reproduced quite well.

```{r}
rbind(
  fit3 |> 
    as_draws() |> 
    subset_draws(c("a","b","sigma")) |> 
    summarise_draws("mean", "mcse_mean", "sd","mcse_sd", "rhat", "ess_bulk") |> 
    mutate(Vx="pchip"),
  fit2 |> 
    as_draws() |> 
    subset_draws(c("a","b","sigma")) |> 
    summarise_draws("mean", "mcse_mean", "sd","mcse_sd", "rhat", "ess_bulk") |> 
    mutate(Vx="spline"),
  fit1 |> 
    as_draws() |> 
    subset_draws(c("a","b","sigma")) |> 
    summarise_draws("mean", "mcse_mean", "sd","mcse_sd", "rhat", "ess_bulk") |> 
    mutate(Vx="true Vx")
)  |> 
  relocate(Vx, .before = 1) |> 
  arrange(variable,Vx)
```

However, the run times with look-up tables are quite a bit longer than using the true expression. The stan implementations are probably far from optimal, even if it is probably not possible to be as fast as evaluating a single small polynomial.

```{r}
rbind(
  fit1$time()$chains |> as_data_frame() |> mutate(model="true"),
  fit2$time()$chains |> as_data_frame() |> mutate(model="spline"),
  fit3$time()$chains |> as_data_frame() |> mutate(model="pchip")
) |> 
  group_by(model) |> 
  summarise(
    avg_sec_per_chain=mean(total)
    )
```
